---
title: "Bayesian Poisson Mixed Model"
author: "Astrid Elmann Hansen"
date: "2026-05-20"
output:
  html_document:
    output_dir: "../html_output"
---
# Setup

```{r setup}
#install.packages("pacman")
pacman::p_load(tidyverse, brms, cmdstanr, bayesplot, posterior)

options(brms.backend = "cmdstanr")

theme_set(theme_minimal())
```

```{r}
sessionInfo()
```


# Load data

```{r}
df <- read_csv('../data/round.csv')
df$condition_FUN <- ifelse(df$condition == "FUN", 1, 0)
```

# Inspect data

```{r}
glimpse(df)
```

# Convert variables

```{r}
df <- df %>%
  arrange(player_id, round) %>%                       # ensure within-player chronology
  group_by(player_id, condition) %>%
  mutate(encounter = row_number()) %>%                # 1st vs 2nd time player saw this condition
  ungroup() %>%
  mutate(
    player_id     = factor(player_id),
    group         = factor(group),
    condition_FUN = factor(condition_FUN),
    encounter     = factor(encounter)
  )
```

# Priors - see 01_H1_bayesian_priors.ipynb for details

```{r}
priors <- c(
  prior(normal(1, 1),   class = "Intercept"), 
  prior(normal(0.2, 0.5), class = "b"),     
  prior(normal(0, 1),   class = "sd")         
)
```


# Fit model

```{r}
fit <- brm(
  formula = tags_in_round ~ condition_FUN + (1 | group/player_id),
  # (1 | group/player_id) expands to (1 | group) + (1 | group:player_id).
  # Players are nested in groups (rounds were played group-wise), so a
  # group-level intercept is part of the design structure regardless of
  # whether σ_group ends up large. With 8 groups and the HalfNormal(0,1)
  # prior on `sd`, σ_group is conservatively regularised toward zero if
  # there is little group-level variance to detect.
  
  data = df,
  
  family = poisson(),
  
  prior = priors,
  
  chains = 4,
  iter = 3000,
  warmup = 1000,
  cores = 4,
  
  backend = "cmdstanr",
  
  seed = 42,
  
  control = list(
    adapt_delta = 0.95
  )
)
```

# Model summary

```{r}
summary(fit)
```

# Diagnostics

## Trace plots

```{r}
plot(fit)
```

## Rhat and ESS

```{r}
# R-hat and ESS from model summary
s <- summary(fit)
s$fixed        # fixed effects with Rhat and ESS
s$random       # random effects (sigma_player AND sigma_group) with Rhat and ESS
```

```{r}
# Extract the two random-intercept SDs explicitly
post <- as_draws_df(fit)

sigma_group  <- post$sd_group__Intercept
sigma_player <- post$`sd_group:player_id__Intercept`

cat("σ_group  (between-group variation):\n")
cat("  Median: ", round(median(sigma_group), 3), "\n")
cat("  95% CrI: [", round(quantile(sigma_group, 0.025), 3),
              ",", round(quantile(sigma_group, 0.975), 3), "]\n\n")

cat("σ_player (between-player variation, within group):\n")
cat("  Median: ", round(median(sigma_player), 3), "\n")
cat("  95% CrI: [", round(quantile(sigma_player, 0.025), 3),
              ",", round(quantile(sigma_player, 0.975), 3), "]\n")
```


```{r}
max_rhat <- max(rhat(fit))
min_ess  <- min(neff_ratio(fit)) * 8000   # convert ratio to absolute ESS

cat("Max R-hat:", round(max_rhat, 3), "\n")
cat("Min ESS:  ", round(min_ess), "\n")
```

# Posterior distributions

```{r}
posterior_summary(fit)
```

# Posterior predictive checks

```{r}
pp_check(fit)
```


# Plot marginal effects

```{r}
plot(conditional_effects(fit))
```

```{r}
post <- as_draws_df(fit)

irr_samples <- exp(post$b_condition_FUN1)

cat("Posterior IRR (FUN vs WIN):\n")
cat("  Median:    ", round(median(irr_samples), 3), "\n")
cat("  95% CrI:  [", round(quantile(irr_samples, 0.025), 3),
                 ",", round(quantile(irr_samples, 0.975), 3), "]\n")
cat("  P(IRR>1): ", round(mean(irr_samples > 1), 4), "\n")
```


```{r}
tibble(irr = irr_samples) |>
  ggplot(aes(x = irr)) +
  geom_histogram(bins = 60, fill = "#DC267F", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_vline(xintercept = median(irr_samples), linetype = "dashed") +
  labs(x = "Incidence Rate Ratio (FUN / WIN)",
       y = "Posterior samples",
       title = paste0("Posterior P(IRR > 1) = ", round(mean(irr_samples > 1), 4))) +
  theme_minimal()
```


# Save model

```{r}
saveRDS(fit, "02_H1_bayes_poisson_model.rds")
```


# Sensitivity check: adding encounter

The pilot showed strong condition-order effects, which is why the main study
counterbalanced WFWF vs FWFW. Counterbalancing should average order effects
out of the H1 estimate, but as a sensitivity check we refit the same model
adding `encounter` (1st vs 2nd time the player encountered this condition)
as a fixed effect. If the condition IRR is stable, the primary H1 result is
robust to the order/encounter dynamic.

```{r}
fit_enc <- brm(
  formula = tags_in_round ~ condition_FUN + encounter + (1 | group/player_id),
  
  data = df,
  
  family = poisson(),
  
  prior = priors,
  
  chains = 4,
  iter = 3000,
  warmup = 1000,
  cores = 4,
  
  backend = "cmdstanr",
  
  seed = 42,
  
  control = list(
    adapt_delta = 0.95
  )
)
```

```{r}
summary(fit_enc)
```

```{r}
# IRR for condition_FUN, marginal over encounter
post_enc <- as_draws_df(fit_enc)
irr_enc  <- exp(post_enc$b_condition_FUN1)

# IRR for encounter (encounter 2 vs encounter 1)
irr_enc_enc <- exp(post_enc$b_encounter2)

cat("── Sensitivity model (with encounter as covariate) ──\n\n")
cat("Condition IRR (FUN vs WIN):\n")
cat("  Median:  ", round(median(irr_enc), 3), "\n")
cat("  95% CrI: [", round(quantile(irr_enc, 0.025), 3),
               ",", round(quantile(irr_enc, 0.975), 3), "]\n")
cat("  P(IRR>1):", round(mean(irr_enc > 1), 4), "\n\n")

cat("Encounter IRR (2nd vs 1st):\n")
cat("  Median:  ", round(median(irr_enc_enc), 3), "\n")
cat("  95% CrI: [", round(quantile(irr_enc_enc, 0.025), 3),
               ",", round(quantile(irr_enc_enc, 0.975), 3), "]\n")
```

```{r}
# Direct comparison with the primary model
irr_primary <- exp(as_draws_df(fit)$b_condition_FUN1)

comparison <- tibble::tribble(
  ~model,           ~median,                ~lower,                              ~upper,
  "Primary",        median(irr_primary),    quantile(irr_primary, 0.025),        quantile(irr_primary, 0.975),
  "+ encounter",    median(irr_enc),        quantile(irr_enc, 0.025),            quantile(irr_enc, 0.975),
)
comparison %>% mutate(across(c(median, lower, upper), ~round(., 3)))
```

```{r}
saveRDS(fit_enc, "02_H1_bayes_poisson_model_encounter.rds")
```


```{r}
# Load model
fit <- readRDS("02_H1_bayes_poisson_model.rds")
```

```{r}
post <- as_draws_df(fit)

irr_samples <- exp(post$b_condition_FUN1)

tibble(irr = irr_samples) |>
  ggplot(aes(x = irr)) +
  geom_histogram(bins = 60, fill = "#DC267F", alpha = 0.7, colour = NA) +
  geom_vline(xintercept = 1,                linetype = "dotted", colour = "#5A5A7A") +
  geom_vline(xintercept = median(irr_samples), linetype = "dashed", colour = "#1C1C2E") +
  labs(x = "Incidence Rate Ratio (FUN / WIN)",
       y = "Posterior samples",
       title = paste0("Posterior P(IRR > 1) > 0.9999")) +
  theme_minimal(base_size = 9) +
  theme(
    plot.background  = element_rect(fill = "#F8F5F0", colour = NA),
    panel.background = element_rect(fill = "#F8F5F0", colour = NA),
    panel.grid.major = element_line(colour = "#E2DDD6"),
    panel.grid.minor = element_blank(),
    axis.text        = element_text(colour = "#5A5A7A"),
    axis.title       = element_text(colour = "#5A5A7A"),
    plot.title       = element_text(colour = "#1C1C2E", face = "bold"),
    axis.line        = element_line(colour = "#E2DDD6"),
    axis.ticks       = element_line(colour = "#E2DDD6")
  )

ggsave("../plots/posterior_irr.png", width = 7, height = 4.5, dpi = 200, bg = "#F8F5F0")
```

# Fit model

```{r}
df <- df %>% filter(group != "E")

fit <- brm(
  formula = tags_in_round ~ condition_FUN + (1 | group/player_id),
  # (1 | group/player_id) expands to (1 | group) + (1 | group:player_id).
  # Players are nested in groups (rounds were played group-wise), so a
  # group-level intercept is part of the design structure regardless of
  # whether σ_group ends up large. With 8 groups and the HalfNormal(0,1)
  # prior on `sd`, σ_group is conservatively regularised toward zero if
  # there is little group-level variance to detect.
  
  data = df,
  
  family = poisson(),
  
  prior = priors,
  
  chains = 4,
  iter = 3000,
  warmup = 1000,
  cores = 4,
  
  backend = "cmdstanr",
  
  seed = 42,
  
  control = list(
    adapt_delta = 0.95
  )
)
```

# Model summary

```{r}
summary(fit)
```

# Diagnostics

## Trace plots

```{r}
plot(fit)
```

## Rhat and ESS

```{r}
# R-hat and ESS from model summary
s <- summary(fit)
s$fixed        # fixed effects with Rhat and ESS
s$random       # random effects (sigma_player AND sigma_group) with Rhat and ESS
```

```{r}
# Extract the two random-intercept SDs explicitly
post <- as_draws_df(fit)

sigma_group  <- post$sd_group__Intercept
sigma_player <- post$`sd_group:player_id__Intercept`

cat("σ_group  (between-group variation):\n")
cat("  Median: ", round(median(sigma_group), 3), "\n")
cat("  95% CrI: [", round(quantile(sigma_group, 0.025), 3),
              ",", round(quantile(sigma_group, 0.975), 3), "]\n\n")

cat("σ_player (between-player variation, within group):\n")
cat("  Median: ", round(median(sigma_player), 3), "\n")
cat("  95% CrI: [", round(quantile(sigma_player, 0.025), 3),
              ",", round(quantile(sigma_player, 0.975), 3), "]\n")
```


```{r}
max_rhat <- max(rhat(fit))
min_ess  <- min(neff_ratio(fit)) * 8000   # convert ratio to absolute ESS

cat("Max R-hat:", round(max_rhat, 3), "\n")
cat("Min ESS:  ", round(min_ess), "\n")
```

# Posterior distributions

```{r}
posterior_summary(fit)
```

# Posterior predictive checks

```{r}
pp_check(fit)
```


# Plot marginal effects

```{r}
plot(conditional_effects(fit))
```

```{r}
post <- as_draws_df(fit)

irr_samples <- exp(post$b_condition_FUN1)

cat("Posterior IRR (FUN vs WIN):\n")
cat("  Median:    ", round(median(irr_samples), 3), "\n")
cat("  95% CrI:  [", round(quantile(irr_samples, 0.025), 3),
                 ",", round(quantile(irr_samples, 0.975), 3), "]\n")
cat("  P(IRR>1): ", round(mean(irr_samples > 1), 4), "\n")
```


```{r}
tibble(irr = irr_samples) |>
  ggplot(aes(x = irr)) +
  geom_histogram(bins = 60, fill = "#DC267F", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_vline(xintercept = median(irr_samples), linetype = "dashed") +
  labs(x = "Incidence Rate Ratio (FUN / WIN)",
       y = "Posterior samples",
       title = paste0("Posterior P(IRR > 1) = ", round(mean(irr_samples > 1), 4))) +
  theme_minimal()
```
