---
title: "H2 — Self-handicap: Wilcoxon tests"
author: "Astrid Elmann Hansen"
date: "2026-05-20"
output:
  html_document:
    output_dir: "../html_output"
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
pacman::p_load(tidyverse, rstatix, ggpubr)
theme_set(theme_minimal())
```

```{r}
sessionInfo()
```

# Load data

```{r}
df <- read_csv('../data/mean_round.csv')
```

All items coded higher = more self-handicap:

* `mean_composite_self_handicap` : mean of the three items
* `mean_ran_max_speed`           : "I ran as fast as possible" (1=very often … 5=never)
* `mean_REVERSED_gave_chance`    : "I gave the catcher a chance on purpose" (reversed)
* `mean_REVERSED_let_close`      : "I let the catcher get near me on purpose" (reversed)

# Inspect

```{r}
df %>%
  select(player_id, condition,
         mean_composite_self_handicap,
         mean_ran_max_speed, mean_REVERSED_gave_chance, mean_REVERSED_let_close) %>%
  glimpse()
```


# Scale reliability

```{r reliability, message=FALSE, warning=FALSE}
items_mat <- df %>%
  select(mean_ran_max_speed, mean_REVERSED_gave_chance, mean_REVERSED_let_close)

cat("N for reliability analysis:", nrow(items_mat), "\n\n")

cat("Inter-item correlations:\n")
print(round(cor(items_mat, use = "complete.obs"), 3))

a <- psych::alpha(items_mat)
cat("\nCronbach's alpha:", round(a$total$raw_alpha, 3), "\n")
cat("95% CI: [",
    round(a$total$raw_alpha - 1.96 * a$total$ase, 3), ",",
    round(a$total$raw_alpha + 1.96 * a$total$ase, 3), "]\n")
```


# Pivot to wide (one row per player)

```{r}
items <- c("mean_composite_self_handicap",
           "mean_ran_max_speed",
           "mean_REVERSED_gave_chance",
           "mean_REVERSED_let_close")

wide <- df %>%
  select(player_id, condition, all_of(items)) %>%
  pivot_wider(names_from  = condition,
              values_from = all_of(items))

glimpse(wide)
```

# Primary: composite — paired Wilcoxon

```{r}
wt_composite <- wilcox.test(
  wide$mean_composite_self_handicap_FUN,
  wide$mean_composite_self_handicap_WIN,
  paired    = TRUE,
  exact     = FALSE,
  conf.int  = TRUE
)
wt_composite

# Effect size r = Z / sqrt(N)
n   <- nrow(wide)
z   <- qnorm(wt_composite$p.value / 2, lower.tail = FALSE)
r   <- z / sqrt(n)

cat("\nEffect size r =", round(r, 3), "\n")
cat("n higher in FUN:", sum(wide$mean_composite_self_handicap_FUN >
                             wide$mean_composite_self_handicap_WIN, na.rm = TRUE), "\n")
cat("n higher in WIN:", sum(wide$mean_composite_self_handicap_FUN <
                             wide$mean_composite_self_handicap_WIN, na.rm = TRUE), "\n")
cat("n tied:         ", sum(wide$mean_composite_self_handicap_FUN ==
                             wide$mean_composite_self_handicap_WIN, na.rm = TRUE), "\n")
```

# Secondary: item-level — Bonferroni corrected (α = 0.05 / 3 ≈ 0.017)

```{r}
item_cols <- c("mean_ran_max_speed",
               "mean_REVERSED_gave_chance",
               "mean_REVERSED_let_close")

item_results <- map_dfr(item_cols, function(col) {
  fun_col <- paste0(col, "_FUN")
  win_col <- paste0(col, "_WIN")
  wt <- wilcox.test(wide[[fun_col]], wide[[win_col]],
                    paired = TRUE, exact = FALSE)
  z_i <- qnorm(wt$p.value / 2, lower.tail = FALSE)
  tibble(
    item      = col,
    W         = wt$statistic,
    p_raw     = wt$p.value,
    r         = z_i / sqrt(nrow(wide)),
    n_fun_hi  = sum(wide[[fun_col]] > wide[[win_col]], na.rm = TRUE),
    n_win_hi  = sum(wide[[fun_col]] < wide[[win_col]], na.rm = TRUE)
  )
})

# Bonferroni correction
item_results <- item_results %>%
  mutate(
    p_bonferroni = p.adjust(p_raw, method = "bonferroni"),
    sig_bonf     = p_bonferroni < 0.05
  )

item_results %>%
  mutate(across(c(p_raw, p_bonferroni), ~formatC(., format="e", digits=2)),
         across(c(W, r), ~round(., 3)))
```

# Descriptive summary

```{r}
df %>%
  group_by(condition) %>%
  summarise(across(all_of(items),
                   list(median = median, mean = mean, sd = sd),
                   .names = "{.col}__{.fn}"),
            .groups = "drop") %>%
  pivot_longer(-condition, names_to = c("item", "stat"), names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(across(c(median, mean, sd), ~round(., 2)))
```

# Plot — composite

```{r fig.width=8, fig.height=4}
p1 <- df %>%
  ggplot(aes(x = condition, y = mean_composite_self_handicap, fill = condition)) +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  geom_jitter(aes(colour = condition), width = 0.08, alpha = 0.5, size = 1.5) +
  scale_fill_manual(values   = c(FUN = "#DC267F", WIN = "#648FFF")) +
  scale_colour_manual(values = c(FUN = "#DC267F", WIN = "#648FFF")) +
  labs(y = "composite self-handicap", x = NULL,
       title = "Composite self-handicap by condition") +
  theme(legend.position = "none")

p2 <- wide %>%
  select(player_id,
         FUN = mean_composite_self_handicap_FUN,
         WIN = mean_composite_self_handicap_WIN) %>%
  pivot_longer(-player_id, names_to = "condition", values_to = "score") %>%
  ggplot(aes(x = condition, y = score)) +
  geom_line(aes(group = player_id), colour = "grey70", alpha = 0.6) +
  geom_point(aes(colour = condition), size = 2, alpha = 0.8) +
  scale_colour_manual(values = c(FUN = "#DC267F", WIN = "#648FFF")) +
  labs(y = NULL, x = NULL,
       title = "Within-player change") +
  theme(legend.position = "none")

ggarrange(p1, p2, ncol = 2)
```

# Plot — individual items

```{r fig.width=10, fig.height=4}
item_labels <- c(
  mean_ran_max_speed          = "Ran at max speed",
  mean_REVERSED_gave_chance   = "Gave tagger a chance",
  mean_REVERSED_let_close     = "Let tagger get close"
)

df %>%
  select(player_id, condition, all_of(item_cols)) %>%
  pivot_longer(all_of(item_cols), names_to = "item", values_to = "score") %>%
  mutate(item = recode(item, !!!item_labels)) %>%
  ggplot(aes(x = condition, y = score, fill = condition)) +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  geom_jitter(aes(colour = condition), width = 0.08, alpha = 0.4, size = 1.2) +
  scale_fill_manual(values   = c(FUN = "#DC267F", WIN = "#648FFF")) +
  scale_colour_manual(values = c(FUN = "#DC267F", WIN = "#648FFF")) +
  facet_wrap(~item, nrow = 1) +
  labs(y = "score (1–5, higher = more self-handicap)", x = NULL,
       title = "Individual items by condition (Bonferroni α = 0.017)") +
  theme(legend.position = "none")
```
